---
title: "Programming Data Visualisation in R"
description: |
  This in-class exercise explores programming data visualisation in R. It introduces the "Grammar of Graphics", ggplot2 for static graphics, ggiraph and plotly for interactive graphics, and tidyverse for the data science workflow.

preview: preview_image.png
author:
  - name: Archie Dolit
    url: https://www.linkedin.com/in/adolit/
    affiliation: School of Computing and Information Systems, Singapore Management University
    affiliation_url: https://scis.smu.edu.sg/
date: 06-26-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
categories:
  - Static Visualisation  
  - Interactive Visualisation
  - R
  - In-class Exercise
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina=3,
                      echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Installing R Packages and Importing Data

## Install and Lauch R Packages

Check, install and launch _ggiraph, plotly, DT_ and _tidyverse_ packages of R

```{r r package}
packages = c('DT', 'ggiraph', 'plotly', 'tidyverse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}
```

## Importing Data

Use read_csv() of readr package to import Exam_data.csv into R

```{r data import}
exam_data <- read_csv("data/Exam_data.csv")
glimpse(exam_data)
summary(exam_data)
```

* Year end examination grades of a cohort of primary 3 students from a local school.

* There are a total of seven attributes. Four of them are categorical data type and the other three are in continuous data type.
  + The categorical attributes are: ID, CLASS, GENDER and RACE.
  + The continuous attributes are: MATHS, ENGLISH and SCIENCE.

# Static Visualisation 

## Comparing Base R Histogram vs _ggplot 2_

### Base R histogram

```{r}
hist(exam_data$MATHS)
```

### _ggplot2_ histogram

```{r}
ggplot(data = exam_data, aes(x=MATHS)) +
  geom_histogram( bins = 10,
                  boundary = 100,
                  color = "black",
                  fill = "grey") +
  ggtitle("Distribution of Maths Score")
```


## Essential Elements in _ggplot2_

### Geometric Objects: _geom_bar_

Plot a bar chart

```{r}
ggplot(data = exam_data,
       aes(x=RACE)) +
  geom_bar()
```

### Geometric Objects: _geom_dotplot_

The width of a dot corresponds to the bin width (or maximum width, depending on the binning algorithm), and dots are stacked, with each dot representing one observation.

```{r}
ggplot(data = exam_data,
       aes(x=MATHS,
           fill = RACE)) +
  geom_dotplot(binwidth = 2.5,
               dotsize = 0.5) +
  scale_y_continuous(NULL,
                     breaks = NULL)
```

### Geometric Objects: _geom_histogram_

geom_histogram() is used to create a simple histogram by using values in MATHS field of exam_data:

 * _bin_ argument was changed to 20 from the defaul value of 30

 * _color_ argument, used to change the outline colour, was set to black

 * _fill_ argument, used to shade the histogram, was set to light blue

```{r}
ggplot(data = exam_data,
       aes(x=MATHS)) + 
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
         
```

### Modifying a geometric object by changing _aes()_

The interior colour of the histogram was changed using the sub-group of _aesthetics_ and _fill_ argument

```{r}
ggplot(data = exam_data,
       aes(x=MATHS,
           fill = GENDER)) + 
  geom_histogram(bins = 20,
                 color = "grey30")
         
```

### Geometric Objects: _geom-density_

geom-density() computes and plots kernel density estimate, which is a smoothed version of the histogram

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()
```

Two kernel density lines by using _colour_ or _fill_ arguments of _aes()_

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```


### Geometric Objects: _geom_boxplot_

Boxplots by using geom_boxplot()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,
           x= GENDER)) +
  geom_boxplot()
```

**Notches** are used in box plots to help visually assess whether the medians of distributions differ. If the notches do not overlap, this is evidence that the medians are different.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

### Combined _geom_ Objects

Plot data points using both _geom_boxplot()_ and _geom_point()_

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) + 
  geom_boxplot() +
  geom_point(position = "jitter",
             size = 0.5)
```

# Interactive Data Visualisation with R - _ggiraph_ methods

## Interactive dotplot

## Tooltip effect with _tooltip_ aesthetic

Interactivity: hovering displays student's ID

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
    ggobj = p,
    width_svg = 6,
    height_svg = 6*0.618
  )
```

##  Hover effect with *data_id* aesthetic

Interactivity: Elements associated with a data_id (i.e CLASS) will be highlighted upon mouse over

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS,
        tooltip = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)
girafe(
    ggobj = p,
    width_svg = 6,
    height_svg = 6*0.618
  )
```

## Styling hover effect

_css codes_ are used to change the highlighting effect

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  ))
```

## Click effect with onclick

```{r}
exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school", as.character(exam_data$ID) )
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(onclick = onclick),
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618)
```
# Interactive Data Visualisation with R - _plotly_ methods

## Interactive scatter plot

```{r}
plot_ly(data = exam_data, 
             x = ~MATHS, 
             y = ~ENGLISH)
```

## Visual Variable

_color_ argument is mapped to a qualitative visual variable (i.e. RACE)

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```

## Changing colour pallete

_colors_ argument is used to change the default colour palette to [ColorBrewel](https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html) colour palette.

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE, 
        colors = "Set1")
```

## Customising colour scheme

```{r}
pal <- c("red", "purple", "blue", "green")
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE, 
        colors = pal)
```

## Customising tooltip

_text_ argument is used to change the default tooltip

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS,
        text = ~paste("Student ID:", ID,
                      "<br>Class:", CLASS),
        color = ~RACE, 
        colors = "Set1")
```

## Working with layout

_layout_ argument is used to change the default tooltip.

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS,
        text = ~paste("Student ID:", ID,     
                      "<br>Class:", CLASS),  
        color = ~RACE, 
        colors = "Set1") %>%
  layout(title = 'English Score versus Maths Score ',
         xaxis = list(range = c(0, 100)),
         yaxis = list(range = c(0, 100)))
```

# Interactive Data Visualisation with R - _ggplotly_ methods

## Interactive scatter plot

Only extra line you need to include in the code chunk is _ggplotly()_

```{r}
p <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(dotsize = 1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```
## Coordinated Multiple Views with plotly

Two scatterplots and places them next to each other side-by-side by using _subplot()_ of **plotly** package

```{r}
p1 <- ggplot(data=exam_data, 
              aes(x = MATHS,
                  y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
p2 <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```


## Coordinated Multiple Views with plotly

To create a coordinated scatterplots, _highlight_key()_ of **plotly package** is used

```{r}
d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```
Click on a data point of one of the scatterplot and see how the corresponding point on the other scatterplot is selected.

Thing to learn from the code chunk:

* _highlight_key()_ simply creates an object of class crosstalk::SharedData.

# Interactive Data Table: DT package
* A wrapper of the JavaScript Library [DataTables](https://datatables.net/)

* Data objects in R can be rendered as HTML tables using the JavaScript library 'DataTables' (typically via R Markdown or Shiny). 

```{r}
DT::datatable(exam_data)
```


## Linked brushing: crosstalk method

Two scatterplots and places them next to each other side-by-side by using _subplot()_ of **plotly** package

```{r}
d <- highlight_key(exam_data)
p <- ggplot(d, aes(ENGLISH, MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
gg <- highlight(ggplotly(p), 
                "plotly_selected")
crosstalk::bscols(gg, DT::datatable(d), widths = 5)
```

Things to learn from the code chunk:

* *highlight()* is a function of **plotly** package. It sets a variety of options for brushing (i.e., highlighting) multiple plots. These options are primarily designed for linking multiple plotly graphs, and may not behave as expected when linking plotly to another htmlwidget package via crosstalk.  In some cases, other htmlwidgets will respect these options, such as persistent selection in leaflet.

* *bscols()* is a helper function of **crosstalk** package.  It makes it easy to put HTML elements side by side.  It can be called directly from the console but is especially designed to work in an R Markdown document.  **Warning:**  This will bring in all of Bootstrap!. 

# Reference
 - [Lesson 7: Programming Data Visualisation in R In-Class Exercise](https://isss608.netlify.app/in-class_ex/in-class_ex07/in-class_ex07-interactiver#1)