---
title: "Visual Detective Assignment"
description: |
  VAST Challenge 2021: Mini-Challenge 2

preview: preview_image.png  
author:
  - name: Archie Dolit
    url: https://www.linkedin.com/in/adolit/
    affiliation: School of Computing and Information Systems, Singapore Management University
    affiliation_url: https://scis.smu.edu.sg/
date: 07-18-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
categories:
  - Visual Detective
  - R
  - Assignment
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina=3,
                      echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

# Overview

# Literature Review

# Exploration and Data Preparation

## Install and Lauch R Packages

```{r r package}
packages = c('ggiraph', 'plotly',
             'sf', 'tmap', 
             'igraph', 'tidygraph', 
             'ggraph', 'visNetwork', 
             'lubridate', 'clock',
             'widyr', 'wordcloud',
             'ggwordcloud', 'DT',
             'textplot', 'hms',
             'tidyverse')

for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

## Importing Data

```{r}
cc <- read_csv("data/aspatial/cc_data.csv")
loyal <- read_csv("data/aspatial/loyalty_data.csv")

head(cc)
head(loyal)
```

# Exploratory Data Analysis

# Mini Challenge 2

## Q1: Most Popular Locations based on Credit Card and Loyalty Data

## Q2: Anomalies in Vehicle, Credit Card and Loyalty Card Data

## Q3: Owners of Credit Card and Loyalty Card

## Q4: Relationships among GASTech Personnel

## Q5: Evidence of Suspicious Activities

## Proposed Visual Analytics RShiny App

# Conclusion

## Reference:

* [ISSS608 Assignment Instructions](https://isss608.netlify.app/assignment.html)

* [VAST Challenge 2021: Mini-Challenge 2](https://vast-challenge.github.io/2021/MC2.html)

* [VAST Challenge 2014: MC2 - Patterns of Life Analysis Benchmark Repository](http://visualdata.wustl.edu/varepository/VAST%20Challenge%202014/challenges/MC2%20-%20Patterns%20of%20Life%20Analysis/)
