---
title: "Visual Detective Assignment Part 4"
description: |
  This assignment attempts to solve the 2021 IEEE Visual Analytics Science and Technology (VAST) Challenge: Mini-Challenge 2 by applying different visual analytics concepts, methods, and techniques with relevant R data visualisation and data analysis packages.

preview: img/preview_image.png  
author:
  - name: Archie Dolit
    url: https://www.linkedin.com/in/adolit/
    affiliation: School of Computing and Information Systems, Singapore Management University
date: 07-25-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
categories:
  - Visual Detective
  - R
  - Assignment
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina=3,
                      echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r r package, echo = FALSE}
packages = c('ggiraph', 'plotly','lobstr',
             'raster','sf', 'tmap', 
             'igraph', 'tidygraph', 
             'ggraph', 'visNetwork', 
             'lubridate', 'clock',
             'widyr', 'wordcloud',
             'ggwordcloud', 'DT',
             'textplot', 'hms',
             'timetk','tidyverse')

for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

```{r import csv, echo = FALSE}
car_data <- read_csv("data/aspatial/car-assignments.csv")
cc_data <- read_csv("data/aspatial/cc_data.csv")
loyalty_data <- read_csv("data/aspatial/loyalty_data.csv")
gps_data <- read_csv("data/aspatial/gps.csv")
```

```{r bgmap, echo = FALSE}
bgmap <- raster("data/Geospatial/abila_map.tif")
# 
# tm_shape(bgmap) +
# tm_rgb(bgmap, r = 1,g = 2,b = 3,
#        alpha = NA,
#        saturation = 1,
#        interpolate = TRUE,
#        max.value = 255)
```

```{r prep car, echo = FALSE}
car_data <- car_data %>%
  #concatenate first and last name
  mutate(FullName = paste(FirstName, LastName, sep = " ")) %>%
  rename(Deparment = CurrentEmploymentType) %>%
  rename(Title = CurrentEmploymentTitle)

car_data$CarID <- as_factor(car_data$CarID)
```

```{r prep cc, echo = FALSE}

#detect and replace Katerina to Katerina's Cafe
cc_data <- cc_data %>%
    mutate(location = ifelse(str_detect(location, "Katerina"), "Katerina's Cafe", location))

#convert to date-time format
cc_data$date <- date_time_parse(cc_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")
cc_data$day <- wday(cc_data$date,
                          label = TRUE,
                          abbr = TRUE)

cc_data$timestamp <- date_time_parse(cc_data$timestamp,
                zone = "",
                format = "%m/%d/%Y %H:%M")

cc_data$hour <- get_hour(cc_data$timestamp)
```

```{r prep loyal, echo = FALSE}

#detect and replace Katerina to Katerina's Cafe
loyalty_data <- loyalty_data %>%
    mutate(location = ifelse(str_detect(location, "Katerina"), "Katerina's Cafe", location))

#convert to date-time format
loyalty_data$date <- date_time_parse(loyalty_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")

loyalty_data$timestamp <- date_time_parse(loyalty_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")

loyalty_data$day <- wday(loyalty_data$timestamp,
                          label = TRUE,
                          abbr = TRUE)
```

```{r prep gps, echo = FALSE}

#rename columns for consistency
gps_data <- gps_data %>%
  rename(timestamp = Timestamp) %>%
  rename(CarID = id)

#convert to date-time format
gps_data$date <- date_time_parse(gps_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")

gps_data$day <- as.factor(wday(gps_data$date,
                          label = TRUE,
                          abbr = TRUE))

gps_data$timestamp <- date_time_parse(gps_data$timestamp,
                zone = "",
                format = "%m/%d/%Y %H:%M:%S")

gps_data$hour <- get_hour(gps_data$timestamp)

#convert to factor data type
gps_data$CarID <- as_factor(gps_data$CarID)

#convert to simple feature 
gps_sf <- st_as_sf(gps_data, 
                   coords = c("long", "lat"),
                       crs= 4326)
```

```{r join cc_loyal, echo = FALSE}

#combine based on date, location, price, exclude day and timestamp
cc_loyalty_data <- full_join(cc_data %>% select(-c("day")),
                             loyalty_data %>% select(-c("day","timestamp")), 
                             by = c("date" = "date", 
                                    "location" = "location", 
                                    "price" = "price"))

#get day of the joint data
cc_loyalty_data$day <- wday(cc_loyalty_data$date,
                          label = TRUE,
                          abbr = TRUE)

#rearrange columns
cc_loyalty_data <- cc_loyalty_data %>%
  select("timestamp", "date", "day", "hour", "location", "price", "last4ccnum", "loyaltynum")
```

```{r join car_gps, echo = FALSE}

#combine based on CarID
car_gps_data <- left_join(gps_data, 
                          car_data %>% select(-c("FirstName", "LastName")),
                          by = "CarID")

car_gps_sf <- left_join(gps_sf,
                        car_data %>% select(-c("FirstName", "LastName")),
                        by = "CarID")
```

# 4. Proposed Solutions

## Q4: Relationships among GASTech Personnel

**Given the data sources provided, identify potential informal or unofficial relationships among GASTech personnel. Provide evidence for these relationships.**


Similar to question 3, identify the POIs by computing the difference of gps timestamp. 

Afterwards, identify who are within 'close contact' of each employee based on the difference of their gps coordinates within the same time period.

This can help establish the relationship of GASTech personnel according to their meetings at the same place and at the same time.


```{r q4 poi network}
gps_poi_network <- car_gps_data %>%
  group_by(CarID) %>%
  mutate(poi_diff = timestamp - lag(timestamp, order_by=CarID)) %>%
  mutate(poi = if_else(poi_diff > 60*5,  TRUE, FALSE)) %>%
  filter(poi == TRUE) %>%
  ungroup() %>%
  mutate(lat_diff = lat - lag(lat, order_by=timestamp))%>%
  mutate(long_diff = long - lag(long, order_by=timestamp)) %>%
  mutate(close_contact = if_else(abs(lat_diff) <=0.001 & abs(long_diff) <=0.001, TRUE, FALSE))%>%
  filter(close_contact == TRUE) %>%
  ungroup()
  
glimpse(gps_poi_network)
```


```{r q4 poi network diagram}

 employee_edges <- gps_poi_network %>%
  group_by(date, hour)%>%
  mutate(from = FullName) %>%
  mutate(to = lead(FullName, order_by = timestamp)) %>%
  ungroup() %>%
  group_by(from,to) %>%
  summarise(weight = n())

 employee_nodes <- gps_poi_network %>%
   select(FullName, Deparment, Title) %>%
   rename(id = FullName) %>%
   rename(group = Deparment) %>%
   distinct()
 
 visNetwork(employee_nodes,
            employee_edges,
            main = "Relationships among GASTech Personnel") %>%
   visIgraphLayout(layout = "layout_with_fr") %>%
   visEdges(arrows = "to", 
            smooth = list(enabled = TRUE, 
                          type = "curvedCW")) %>%
   visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
   visLegend() %>%
   visLayout(randomSeed = 123)
```

The network diagram shows the 'official' relationship of employees based on their respective departments. It also show 'unofficial' relationship based on the number of their interactions.

From the  network diagram, it can be seen that _Isande Barrasca_ , a Drill Technician from the Engineering Department is an outlier.
His only close contact to the rest of employees is _Hideki Cocinaro_ , a Site Controller from the Security Department.

Similarly, _Sten Sanjorge Jr._ , IT Technician from the Information Technology Department, have minimal interactions with other employees and seems not well connected within the company.


The heatmap below visualizes the number of interactions between employees.

```{r q4 employee interaction, fig.height=8}
employee_interact1 <- full_join(employee_edges, 
                             employee_nodes, 
                             by = (c("from" = "id"))) %>%
  rename(SenderDepartment = group) %>%
  rename(SenderTitle = Title)

employee_interact2 <- full_join(employee_interact1, 
                             employee_nodes, 
                             by = (c("to" = "id"))) %>%
  rename(ReceiverDepartment = group) %>%
  rename(ReceiverTitle = Title) %>%
  rename(Sender = from) %>%
  rename(Receiver = to)

employee_interaction  <- ggplot(data = employee_interact2,
                         aes(x=Sender, y=Receiver,
                                   fill = weight,
                                   text = paste("Sender :", Sender,"\n",
                                                "Sender Department:", SenderDepartment, "\n",
                                                "Sender Title:", SenderTitle, "\n",
                                                "\n",
                                                "Receiver", Receiver,"\n",
                                                "Receiver Department:", ReceiverDepartment, "\n",
                                                "Receiver Title:", ReceiverTitle, "\n",
                                                "\n",
                                                "Number of Meetings", weight))) +
  geom_tile()+
  scale_fill_gradient(low = "lightsteelblue1", high = "royalblue4") +
  ggtitle("GAStech Personnel Relationship based on number of Interactions") +
  labs(x = "Sender Employee", y = "Receiver Employee") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))
  
ggplotly(employee_interaction, tooltip = "text")
```

The highest number of meetings among GAStech Personnel are truck drivers with 23 interactions. Employee names are set to _NA_ since the CarID is not identified.
The second highest number of meetings is from _Bertrand Ovan_, Group Manager of Facilities department with 14 meetings.  
The third highest number of meetings is from _Ingrid Barranco_, SVP/CFO from Executive department with 12 meetings.


Click [**HERE**](https://adolit-vaa.netlify.app/posts/2021-07-26-assignment-5/) to view the _Visual Detective Assignment Part 5_.