---
title: "Network Data Visualisation and Analysis with R"
description: |
  This in-class exercise explores Network Data Visualisation and Analysis. It visualises the GASTech emails from VAST Challenge 2021: [Mini-Challenge 1](https://vast-challenge.github.io/2021/MC1.html) using dplyr, lubridate, tidygraph, ggraph, and VisNetwork packages. Additionally, the cc_data.csv from VAST Challenge 2021: [Mini-Challenge 2](https://vast-challenge.github.io/2021/MC2.html) was used to create Bipartitle Network.

preview: preview_image.png
author:
  - name: Archie Dolit
    url: https://www.linkedin.com/in/adolit/
    affiliation: School of Computing and Information Systems, Singapore Management University
    affiliation_url: https://scis.smu.edu.sg/
date: 07-10-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
categories:
  - Network Data
  - R
  - In-class Exercise
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina=3,
                      echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Install and Lauch R Packages

```{r r package}
packages = c('igraph', 'tidygraph', 
             'ggraph', 'visNetwork', 
             'lubridate', 'clock',
             'tidyverse')
for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```


## Import Network Data

Import GAStech_email_node.csv and GAStech_email_edges.csv into RStudio environment by using *read_csv()* of **readr** package

```{r import}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```


## Review Imported Data

```{r review}
glimpse(GAStech_edges)
```
From the output report of GAStech_edges, SentDate is treated as "Character"" data type instead of date data type. The data type of SentDate field needs to change back to "Date" data type.

## Data Wrangling

```{r wrangling}
GAStech_edges$SentDate  = dmy(GAStech_edges$SentDate)
GAStech_edges$Weekday = wday(GAStech_edges$SentDate, 
                             label = TRUE, 
                             abbr = FALSE)
```

* both *dmy()* and *wday()* are functions of **lubridate** package.  [lubridate](cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) is an R package that makes it easier to work with dates and times.  

* *dmy()* transforms the SentDate to Date data type.

* *wday()* returns the day of the week as a decimal number or an ordered factor if label is TRUE.  The argument abbr is FALSE keep the daya spells in full, i.e. Monday.  The function will create a new column in the data.frame i.e. Weekday and the output of *wday()* will save in this newly created field.

* the values in the *Weekday* field are in ordinal scale. 

## Review Revised Date Fields

```{r review revised}
glimpse(GAStech_edges)
```

## Wrangling Attributes

GAStech_edges data.frame consists of individual e-mail flow records, need to aggregate the individual by date, senders, receivers, main subject and day of the week

```{r wrangling attributes}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

* four functions from **dplyr** package are used.  They are: *filter()*, *group()*, *summarise()*, and *ungroup()*.  

* The output data.frame is called **GAStech_edges_aggregated**. 

* A new field called *Weight* has been added in GAStech_edges_aggregated.

## Review Revised Edges Files

```{r review edges}
glimpse(GAStech_edges_aggregated)
```

## Using tbl_graph() to build tidygraph data model.

Use *tbl_graph()* of **tinygraph** package to build an tidygraph's network graph data.frame.
Check the reference guide of *tbl_graph()* [here](https://tidygraph.data-imaginist.com/reference/tbl_graph.html)

```{r Code Chuck NAME}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)
```

## Review tidygraph's graph object

```{r review tidygraph}
GAStech_graph
```
- The output above reveals that *GAStech_graph* is a tbl_graph object with 54 nodes and 1456 edges. 

- The command also prints the first six rows of "Node Data" and the first three of "Edge Data". 

- It states that the Node Data is **active**. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time. 


## Change Active Object

```{r active object}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```
The nodes tibble data frame is activated by default, but you can change which tibble data frame is active with the *activate()* function. Thus, if we wanted to rearrange the rows in the edges tibble to list those with the highest "weight" first, we could use *activate()* and then *arrange()*.  

Check the reference guide of *activate()* [here](https://tidygraph.data-imaginist.com/reference/activate.html).

## Plot Basic Network Graph

The code chunk below uses [*ggraph()*](https://ggraph.data-imaginist.com/reference/ggraph.html), [*geom-edge_link()*](https://ggraph.data-imaginist.com/reference/geom_edge_link.html) and [*geom_node_point()*](https://ggraph.data-imaginist.com/reference/geom_node_point.html) to plot a network graph by using *GAStech_graph*.


```{r basic network}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```
The basic plotting function is *ggraph()*, which takes the data to be used for the graph and the type of layout desired. Both of the arguments for *ggraph()* are built around igraph. Therefore, *ggraph()* can use either an igraph object or a tbl_graph object.


## Change Default Network Graph Theme

```{r network theme}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())
g + theme_graph()
```
* **ggraph** introduces a special ggplot theme that provides better defaults for network graphs than the normal ggplot defaults.  *theme_graph()*, besides removing axes, grids, and border, changes the font to Arial Narrow (this can be overridden).  

* The ggraph theme can be set for a series of plots with the *set_graph_style()* command run before the graphs are plotted or by using *theme_graph()* in the individual plots. 

## Change Color of the Plot

```{r network color}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))
g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

## Fruchterman and Reingold Layout

```{r network fr}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())
g + theme_graph()
```

* *layout* argument is used to define the layout to be used. 

## Modify Network Nodes

Colour each node by referring to their respective departments.

```{r modify nodes}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))
g + theme_graph()
```

* *geom_node_point* is equivalent in functionality to *geo_point* of **ggplot2**.  It allows for simple plotting of nodes in different shapes, colours and sizes.


## Modify Network Edges

The thickness of the edges will be mapped with the *Weight* variable.

```{r modify network}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)
g + theme_graph()
```
- *geom_edge_link* draws edges in the simplest way - as straight lines between the start and end nodes.  But, it can do more that that.  In the example above, argument *width* is used to map the width of the line in proportional to the Weight attribute and argument alpha is used to introduce opacity on the line.    

## Working with *facet_edges()*

In the code chunk below, [*facet_edges()*](https://ggraph.data-imaginist.com/reference/facet_edges.html) is used.

```{r facet_edges}
set_graph_style()
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
g + facet_edges(~Weekday)
```

The code chunk below uses *theme()* to change the position of the legend.

```{r facet_edges theme}
set_graph_style()
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```
Add frame to each graph

```{r facet_edges frame}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```
## Working with facet_nodes()

In the code chunck below, [*facet_nodes()*](https://ggraph.data-imaginist.com/reference/facet_nodes.html) is used

```{r face_nodes}
set_graph_style()
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## Computing centrality indices

Centrality measures are a collection of statistical indices use to describe the relative important of the actors are to a network.  There are four well-known centrality measures, namely: degree, betweenness, closeness and eigenvector.


```{r betweeness}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

* *mutate()* of **dplyr** is used to perform the computation.

* the algorithm used, on the other hand, is the *centrality_betweenness()* of **tidygraph**. 

## Visualising network metrics

From **ggraph v2.0** onwards tidygraph algorithms such as centrality measures can be accessed directly in ggraph calls.  This means that it is no longer necessary to precompute and store derived node and edge centrality measures on the graph in order to use them in a plot.

```{r network metrics}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```
## Visualising Community

tidygraph package inherits many of the community detection algorithms imbedded into igraph and makes them available to us, including *Edge-betweenness (group_edge_betweenness)*, *Leading eigenvector (group_leading_eigen)*, *Fast-greedy (group_fast_greedy)*, *Louvain (group_louvain)*, *Walktrap (group_walktrap)*, *Label propagation (group_label_prop)*, *InfoMAP (group_infomap)*, *Spinglass (group_spinglass)*, and *Optimal (group_optimal)*. Some community algorithms are designed to take into account direction or weight, while others ignore it.  Use this [link](https://tidygraph.data-imaginist.com/reference/group_graph.html) to find out more about community detection functions provided by tidygraph. 



```{r Visualising Community}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))  
g + theme_graph()
```

## Interactive Network Graph with visNetwork

## Data Preparation
```{r VisNet Data Preparation}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

* *visNetwork()* function uses a nodes list and edges list to create an interactive graph. 
* The nodes list must include an "id" column, and the edge list must have "from" and "to" columns. 
* The function also plots the labels for the nodes, using the names of the actors from the "label" column in the node list. 


## Plotting the first interactive network graph

```{r interactive network}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```
## Working with layout

In the code chunk below, Fruchterman and Reingold layout is used.

```{r fr interactive}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 
```

## Working with visual attributes - Nodes

visNetwork() looks for a field called "group" in the nodes object and colour the nodes according to the values of the group field.  

The code chunk below rename Department field to group.

```{r rename group}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 
```

When we rerun the code chunk below, visNetwork shades the nodes by assigning unique colour to each category in the *group* field.

```{r color group}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123) 
```

In the code run below *visEdges()* is used to symbolise the edges.  
* The argument *arrows* is used to define where to place the arrow.  
* The *smooth* argument is used to plot the edges using a smooth curve

```{r arrow smooth}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## Interactivity

In the code chunk below, *visOptions()* is used to incorporate interactivity features in the data visualisation.  
* The argument *highlightNearest* highlights nearest when clicking a node.  
* The argument *nodesIdSelection* adds an id node selection creating an HTML select element. 

```{r interactivity}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

## Bipartitle Network

Transform a transaction data set into a network graph by using tidygraph.  The data we are going to use is the *cc_data.csv* data from VAST Challenge 2021: [Mini-Challenge 2](https://vast-challenge.github.io/2021/MC2.html)

## Import Data

Parse the data into R Studio by using *read_csv()*

```{r parse cc data}
cc_data <- read_csv("data/cc_data.csv")
```

Review imported tibble data.frame

```{r glimpse cc data}
glimpse(cc_data)
```
Notice that the *timestamp* field is in character format and not in date-time format.  

## Transform Data

* Transform values in *timestamp* field into correct date-time format
* Transform values in *last4ccnum* field to character data type
* Derive a day-of-month field from timestamp
* Derive a hour-of-day field from timestamp

```{r transform cc data}
cc_data$timestamp <-  date_time_parse(cc_data$timestamp,
                                      zone = "",
                                      format = "%m/%d/%Y %H:%M")
cc_data$last4ccnum <- as.character(cc_data$last4ccnum)
cc_data$Day  = get_day(cc_data$timestamp)
cc_data$Hour = get_hour(cc_data$timestamp)
```

## Create Nodes List

* Get the distinct card users from the “last4ccnum” column.
* Rename the column with the "last4ccnum" as “label” to adopt the vocabulary used by network analysis packages.
* Repeat the same steps for "location" column.
* Join both the sources and destinations data frame

```{r node cc data}
sources <- cc_data %>%
  distinct(last4ccnum) %>%
  rename(label = last4ccnum)

destinations <- cc_data %>%
  distinct(location) %>%
  rename(label = location)
```

To create a single dataframe with a column with the unique users and locations, *full_join()* is used.  T
This is because we want to include all unique places from both the sources and the destinations of ccdata.

```{r join cc data}
cc_nodes <- full_join(sources, 
                   destinations, 
                   by = "label")
```

Use *rowid_to_column()* to add an “id” column to the nodes data frame

```{r rowid cc data}
cc_nodes <- cc_nodes %>% 
  rowid_to_column("id")
```

## Create Edges List

To create an edge list, we need to deal with two ID columns instead of one.
We also want to create a weight column that will note the amount of letters sent between each set of nodes.

```{r edge cc data}
edges <- cc_data %>%  
  group_by(last4ccnum, location, Day, Hour) %>%
  summarise(weight = n()) %>% 
  ungroup()

edges
```

## Tidy Edges List
Like the node list, cc_edges now has the basic form that we want, but we again have the problem that the “source” and “destination” columns contain labels rather than IDs.  

```{r tidy edge cc data}
cc_edges <- edges %>% 
  left_join(cc_nodes, 
            by = c("last4ccnum" = "label")) %>% 
  rename(from = id)

cc_edges <- cc_edges %>% 
  left_join(cc_nodes, 
            by = c("location" = "label")) %>% 
  rename(to = id)
```

Reorder the columns to bring “from” and “to” to the left of the data frame

```{r reorder edge cc data}
cc_edges <- select(cc_edges, from, to, 
                   Day, Hour, weight)
cc_edges
```


## Building tidygraph network graph data object

*tbl_graph()* of **tidygraph** package is used to build a network object by using a node list and an edge list.

```{r tidygraph cc data}
cc_graph <- tbl_graph(nodes = cc_nodes, 
                      edges = cc_edges, 
                      directed = FALSE)

cc_graph
```

## Visualising the network

Network graph created by using cc_graph network object

```{r network graph cc data}
ggraph(cc_graph, 
       layout = "lgl") +
  geom_edge_link(aes()) +
  geom_node_point(aes()) +
  theme_graph()
```

## Reference:
* [Lesson 9: Network Data Visualisation and Analysis In-Class Exercise](https://isss608.netlify.app/in-class_ex/in-class_ex09/in-class_ex09-networkvis#1)