---
title: "Visual Detective Assignment Part 3"
description: |
  This assignment attempts to solve the 2021 IEEE Visual Analytics Science and Technology (VAST) Challenge: Mini-Challenge 2 by applying different visual analytics concepts, methods, and techniques with relevant R data visualisation and data analysis packages.

preview: img/preview_image.png  
author:
  - name: Archie Dolit
    url: https://www.linkedin.com/in/adolit/
    affiliation: School of Computing and Information Systems, Singapore Management University
date: 07-25-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
categories:
  - Visual Detective
  - R
  - Assignment
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina=3,
                      echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r r package, echo = FALSE}
packages = c('ggiraph', 'plotly','lobstr',
             'raster','sf', 'tmap', 
             'igraph', 'tidygraph', 
             'ggraph', 'visNetwork', 
             'lubridate', 'clock',
             'widyr', 'wordcloud',
             'ggwordcloud', 'DT',
             'textplot', 'hms',
             'timetk','tidyverse')

for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

```{r import csv, echo = FALSE}
car_data <- read_csv("data/aspatial/car-assignments.csv")
cc_data <- read_csv("data/aspatial/cc_data.csv")
loyalty_data <- read_csv("data/aspatial/loyalty_data.csv")
gps_data <- read_csv("data/aspatial/gps.csv")
```

```{r bgmap, echo = FALSE}
bgmap <- raster("data/Geospatial/abila_map.tif")
# 
# tm_shape(bgmap) +
# tm_rgb(bgmap, r = 1,g = 2,b = 3,
#        alpha = NA,
#        saturation = 1,
#        interpolate = TRUE,
#        max.value = 255)
```

```{r prep car, echo = FALSE}
car_data <- car_data %>%
  #concatenate first and last name
  mutate(FullName = paste(FirstName, LastName, sep = " ")) %>%
  rename(Deparment = CurrentEmploymentType) %>%
  rename(Title = CurrentEmploymentTitle)

car_data$CarID <- as_factor(car_data$CarID)
```

```{r prep cc, echo = FALSE}

#detect and replace Katerina to Katerina's Cafe
cc_data <- cc_data %>%
    mutate(location = ifelse(str_detect(location, "Katerina"), "Katerina's Cafe", location))

#convert to date-time format
cc_data$date <- date_time_parse(cc_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")
cc_data$day <- wday(cc_data$date,
                          label = TRUE,
                          abbr = TRUE)

cc_data$timestamp <- date_time_parse(cc_data$timestamp,
                zone = "",
                format = "%m/%d/%Y %H:%M")

cc_data$hour <- get_hour(cc_data$timestamp)
```

```{r prep loyal, echo = FALSE}

#detect and replace Katerina to Katerina's Cafe
loyalty_data <- loyalty_data %>%
    mutate(location = ifelse(str_detect(location, "Katerina"), "Katerina's Cafe", location))

#convert to date-time format
loyalty_data$date <- date_time_parse(loyalty_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")

loyalty_data$timestamp <- date_time_parse(loyalty_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")

loyalty_data$day <- wday(loyalty_data$timestamp,
                          label = TRUE,
                          abbr = TRUE)
```

```{r prep gps, echo = FALSE}

#rename columns for consistency
gps_data <- gps_data %>%
  rename(timestamp = Timestamp) %>%
  rename(CarID = id)

#convert to date-time format
gps_data$date <- date_time_parse(gps_data$timestamp,
                zone = "",
                format = "%m/%d/%Y")

gps_data$day <- as.factor(wday(gps_data$date,
                          label = TRUE,
                          abbr = TRUE))

gps_data$timestamp <- date_time_parse(gps_data$timestamp,
                zone = "",
                format = "%m/%d/%Y %H:%M:%S")

gps_data$hour <- get_hour(gps_data$timestamp)

#convert to factor data type
gps_data$CarID <- as_factor(gps_data$CarID)

#convert to simple feature 
gps_sf <- st_as_sf(gps_data, 
                   coords = c("long", "lat"),
                       crs= 4326)
```

```{r join cc_loyal, echo = FALSE}

#combine based on date, location, price, exclude day and timestamp
cc_loyalty_data <- full_join(cc_data %>% select(-c("day")),
                             loyalty_data %>% select(-c("day","timestamp")), 
                             by = c("date" = "date", 
                                    "location" = "location", 
                                    "price" = "price"))

#get day of the joint data
cc_loyalty_data$day <- wday(cc_loyalty_data$date,
                          label = TRUE,
                          abbr = TRUE)

#rearrange columns
cc_loyalty_data <- cc_loyalty_data %>%
  select("timestamp", "date", "day", "hour", "location", "price", "last4ccnum", "loyaltynum")
```

```{r join car_gps, echo = FALSE}

#combine based on CarID
car_gps_data <- left_join(gps_data, 
                          car_data %>% select(-c("FirstName", "LastName")),
                          by = "CarID")

car_gps_sf <- left_join(gps_sf,
                        car_data %>% select(-c("FirstName", "LastName")),
                        by = "CarID")
```

# 4. Proposed Solutions

## Q3: Owners of Credit Card and Loyalty Card


**Can you infer the owners of each credit card and loyalty card? What is your evidence? Where are there uncertainties in your method? Where are there uncertainties in the data?**

It is challenging to infer the owners of the credit card and loyalty card since there is no data field to join the credit card and gps data. 

One possible approach to plot the gps path and check it against timestamp of the credit card transactions. 
This approach assumes that the person driving the car is the same person making the credit card transaction.
Additionally, it assumes that the gps coordinates, timestamp as well as the credit location and timestamp are accurate.

To implement this proposed approach, it assumes the geospatial tracking software installed in the employees car will stop tracking if the vehicle is not moving.
Borrowing from concept of 'Point of Interest' (POI) from Virginia Tech, POI is identified if the employee stops for more than 5 minutes.


Identify the POIs by computing the difference of gps timestamp. If the difference is greater than 5 minutes, it will be set to poi = TRUE.

```{r q3 poi}
gps_poi_sf <- car_gps_sf %>%
  group_by(CarID) %>%
  mutate(diff = timestamp - lag(timestamp, order_by=CarID)) %>%
  mutate(poi = if_else(diff > 60*5,  TRUE, FALSE)) %>%
  filter(poi == TRUE) %>%
  ungroup() 

glimpse(gps_poi_sf)
```


Plot the POIs on the backgroup tourist map and identity the employee locations.
From the  _Combined Credit Card & Loyalty Transactions by Hour of Day_ heatmap of Question 1, the earliest transaction is around 3:00 AM from Kronos Mart while the last transaction is 10:00 PM from Hippokampos. This information can be used to limit the number of POIs.


```{r q3 poi points, fig.height=40}
gps_poi_points <- gps_poi_sf %>%
  filter(hour >= 2 & hour <= 23) %>%
  select(timestamp,
         CarID,
         Deparment,
         Title,
         FullName)

tmap_mode("plot")
tm_shape(bgmap) +
  tm_rgb(bgmap, r = 1,g = 2,b = 3,
       alpha = NA,
       saturation = 1,
       interpolate = TRUE,
       max.value = 255) +
  tm_shape(gps_poi_points) +
  tm_dots(col = 'red', border.col = 'black', size = 1, alpha = 0.5, jitter = .8) +
  tm_facets(by = "FullName", ncol = 2)
```


Alternatively, plot the gps path of each employee.


```{r q3 gps path, fig.height=40}
gps_path <- car_gps_sf %>%
  group_by(CarID) %>%
  summarize(m = mean(timestamp),
            do_union=FALSE) %>%
  st_cast("LINESTRING")

tmap_mode("plot")
tm_shape(bgmap) +
  tm_rgb(bgmap, r = 1,g = 2,b = 3,
       alpha = NA,
       saturation = 1,
       interpolate = TRUE,
       max.value = 255) +
  tm_shape(gps_path) +
  tm_lines() +
  tm_facets(by = "CarID",  ncol = 2)
```



Create an interactive data table based on the joint credit card and infer the owner based on the POI plot.


```{r q3 table cc data}

cc_owner <- cc_data %>%
  select(timestamp, location, last4ccnum)

DT::datatable(cc_owner)
```

Finally, manually map the credit card transaction purchases timestamp against the POI map. This approach is very time consuming since it entails manual effort.
As for most cases, the POI map may show certain points around the vicinity of the location but does not have any corresponding credit card purchases.


The owner of the loyalty card will be known after the credit card owner has been identified. 
As noted earlier, there is no one-to-one correspondence between the credit and loyalty card. 

The heatmap below visualizes the number of transactions between specific credit card and loyalty card to know the highest likelihood of the loyalty card owner.

```{r count cc loyal,fig.height=8}
cc_loyal_count <- cc_loyalty_data %>%
  group_by(last4ccnum,loyaltynum) %>%
  summarise(count=n())
  
cc_loyal_correlate <- ggplot(data = cc_loyal_count,
                               aes(x=loyaltynum, y=as.factor(last4ccnum),
                                   fill = count,
                                   text = paste("Last 4 Credit Card Number :", last4ccnum,"\n",
                                                "Loyalty Card Number:", loyaltynum,"\n",
                                                "Number of transactions :", count))) +
  geom_tile()+
  scale_fill_gradient(low = "lightsteelblue1", high = "royalblue4") +
  ggtitle("Correlation of Credit Card & Loyalty Card by Number of Transactions") +
  labs(x = "Loyalty Card Number",y = "Credit Card Number") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90))

ggplotly(cc_loyal_correlate, tooltip = "text")
```

From the heatmap, it can seen that certain combination of credit card number and loyalty card are most often used together.
Example of which are _6901 and L9363_ with 28 transactions; _7117 and L6417_ also with 28 transactions.

Nevertheless, even for credit card 6901, there are still _9 transactions_ when loyalty card was not used.
Similarly, credit card 7117 has _3 transactions_ when loyalty card was not used.

Click [**HERE**](https://adolit-vaa.netlify.app/posts/2021-07-26-assignment-4/) to view the _Visual Detective Assignment Part 4_.