---
title: "Programming GeoVisual Analytics with R"
description: |
  This in-class exercise explores Analysis of Spatial Data. Specifically, it visualises the movement data from VAST Challenge 2021: [Mini-Challenge 2](https://vast-challenge.github.io/2021/MC2.html) using sf, raster, readr, clock and tmap packages.

preview: preview_image.png
author:
  - name: Archie Dolit
    url: https://www.linkedin.com/in/adolit/
    affiliation: School of Computing and Information Systems, Singapore Management University
    affiliation_url: https://scis.smu.edu.sg/
date: 07-03-2021
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
categories:
  - GeoSpatial
  - R
  - In-class Exercise
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina=3,
                      echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Install and Lauch R Packages

```{r r package}
packages = c('raster', 'sf', 
             'tmap', 'clock', 
             'tidyverse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}
```

## Import Raster file
```{r import}
bgmap <- raster("data/Geospatial/MC2-tourist.tif")
bgmap
```
## Plot Raster Layer

In general, [tm_raster()](https://cran.r-project.org/web/packages/tmap/tmap.pdf) will be used to plot a raster layer by using tmap package.

```{r raster}
tmap_mode("plot")
tm_shape(bgmap) +
    tm_raster(bgmap,
              legend.show = FALSE)
```
However, *bgmap* layer is a three bands false colour image.  Hence, *tm_rgb()* is used instead.

```{r rgb raster}
tm_shape(bgmap) +
tm_rgb(bgmap, r = 1,g = 2,b = 3,
       alpha = NA,
       saturation = 1,
       interpolate = TRUE,
       max.value = 255)
```
## Import Vector GIS Data File

*Abila* GIS data layer is in [ESRI shapefile](https://en.wikipedia.org/wiki/Shapefile) format.  It is in vector data model and the feature class is line.

Using [st_read()](https://r-spatial.github.io/sf/reference/st_read.html) of sf package, import *Abila* shapefile into R.

```{r gis import}
Abila_st <- st_read(dsn = "data/Geospatial",
                    layer = "Abila")
```

## Import Aspatial Data

Using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) of **readr** package, import *gps.csv* into R.

```{r csv import}
gps <- read_csv("data/aspatial/gps.csv")
glimpse(gps)
```
## Convert Date-Time Field

[*data-time_parse()*](https://clock.r-lib.org/reference/date-time-parse.html) of **clock** package is used to convert *Timestamp* filed from *Character* data type to *date-time* (i.e. dttm) format.

```{r convert date}
gps$Timestamp <- date_time_parse(gps$Timestamp,
                                 zone = "",
                                 format = "%m/%d/%Y %H:%M:%S")

gps$day <- as.factor(get_day(gps$Timestamp))
```

[*as_factor()*](https://forcats.tidyverse.org/reference/as_factor.html) of [**forcats**](https://forcats.tidyverse.org/index.html) package is used to convert values in id field from numerical to factor data type.

```{r factor id}
gps$id <- as_factor(gps$id)

gps
```

## Convert Aspatial Data into a Simple Feature Data

Converts *gps* data frame into a simple feature data frame by using [*st_as_sf()*](https://r-spatial.github.io/sf/reference/st_as_sf.html) of **sf** packages 

The *coords* argument requires you to provide the column name of the x-coordinates (i.e. long) first then followed by the column name of the y-coordinates (i.e. lat).

The *crs* argument required you to provide the coordinates system in epsg format.  [EPSG: 4326](https://epsg.io/4326) is wgs84 Geographic Coordinate System.  You can search for other country's epsg code by referring to [epsg.io](https://epsg.io/).

```{r covert aspatial}
gps_sf <- st_as_sf(gps, 
                   coords = c("long", "lat"),
                       crs= 4326)
gps_sf
```

## Create Movement Path from GPS Points

Joins the gps points into movement paths by using the drivers' IDs as unique identifiers

```{r gps movement}
gps_path <- gps_sf %>%
  group_by(id, day) %>%
  summarize(m = mean(Timestamp), 
            do_union=FALSE) %>%
  st_cast("LINESTRING")

gps_path
```


## Plot the gps path

Plot the gps path of driver ID 1 onto the background tourist map

```{r gps path}
gps_path_selected <- gps_path %>%
  filter(id==1)
tmap_mode("view")
tm_shape(bgmap) +
  tm_rgb(bgmap, r = 1,g = 2,b = 3,
       alpha = NA,
       saturation = 1,
       interpolate = TRUE,
       max.value = 255) +
  tm_shape(gps_path_selected) +
  tm_lines()
```
## Reference:
 - [Lesson 8: Programming GeoVisual Analytics with R In-Class Exercise](https://isss608.netlify.app/in-class_ex/in-class_ex08/in-class_ex08-movementvis#1)